%%%%%%%% ICML 2021 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%

\documentclass{article}

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{booktabs}  % for professional tables
\usepackage{xcolor}
% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{icml2021} with \usepackage[nohyperref]{icml2021} above.
\usepackage{hyperref}

% Attempt to make hyperref and algorithmic work together better:
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% For importing packages and defining commands
\input{macro}

% Use the following line for the initial blind version submitted for review:
% \usepackage{icml2021}

% If accepted, instead use the following line for the camera-ready submission:
\usepackage[accepted]{icml2021}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{A Policy Gradient Algorithm for Learning to Learn in Multiagent Reinforcement Learning}

% Comment for the first time, but uncomment for second time with commenting out appendix
% to only have main paper
% \nofiles

\begin{document}

\twocolumn[
\icmltitle{A Policy Gradient Algorithm for Learning to Learn \\in Multiagent Reinforcement Learning}

% It is OKAY to include author information, even for blind
% submissions: the style file will automatically remove it for you
% unless you've provided the [accepted] option to the icml2021
% package.

% List of affiliations: The first argument should be a (short)
% identifier you will use later to specify author affiliations
% Academic affiliations should list Department, University, City, Region, Country
% Industry affiliations should list Company, City, Region, Country

% You can specify symbols, otherwise they are numbered in order.
% Ideally, you should not use this facility. Affiliations will be numbered
% in order of appearance and this is the preferred way.
\icmlsetsymbol{equal}{*}

\begin{icmlauthorlist}
\icmlauthor{Dong-Ki Kim}{mit,mitibm}
\icmlauthor{Miao Liu}{mitibm,ibm}
\icmlauthor{Matthew Riemer}{mitibm,ibm}
\icmlauthor{Chuangchuang Sun}{mit,mitibm}
\icmlauthor{Marwa Abdulhai}{mit,mitibm}
\\
\icmlauthor{Golnaz Habibi}{mit,mitibm}
\icmlauthor{Sebastian Lopez-Cot}{mit,mitibm}
\icmlauthor{Gerald Tesauro}{mitibm,ibm}
\icmlauthor{Jonathan P. How}{mit,mitibm}
\end{icmlauthorlist}

\icmlaffiliation{mit}{MIT-LIDS}
\icmlaffiliation{ibm}{IBM-Research}
\icmlaffiliation{mitibm}{MIT-IBM Watson AI Lab}

\icmlcorrespondingauthor{Dong-Ki Kim}{dkkim93@mit.edu}
\icmlcorrespondingauthor{Miao Liu}{miao.liu1@ibm.com}

% You may provide any keywords that you
% find helpful for describing your paper; these are used to populate
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{Multiagent reinforcement learning, Meta-learning, Non-stationarity}

\vskip 0.3in
]

% this must go after the closing bracket ] following \twocolumn[ ...

% This command actually creates the footnote in the first column
% listing the affiliations and the copyright notice.
% The command takes one argument, which is text to display at the start of the footnote.
% The \icmlEqualContribution command is standard text for equal contribution.
% Remove it (just {}) if you do not need this facility.

\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
%\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.

% For reducing spacing between align math
\setlength{\abovedisplayskip}{3.5pt}
\setlength{\belowdisplayskip}{3.5pt}
\setlength{\abovedisplayshortskip}{3.5pt}
\setlength{\belowdisplayshortskip}{3.5pt}

\begin{abstract}
A fundamental challenge in multiagent reinforcement learning is to learn beneficial behaviors in a shared environment with other simultaneously learning agents. In particular, each agent perceives the environment as effectively non-stationary due to the changing policies of other agents. Moreover, each agent is itself constantly learning, leading to natural non-stationarity in the distribution of experiences encountered. In this paper, we propose a novel meta-multiagent policy gradient theorem that directly accounts for the non-stationary policy dynamics inherent to multiagent learning settings. This is achieved by modeling our gradient updates to consider both an agent’s own non-stationary policy dynamics and the non-stationary policy dynamics of other agents in the environment. We show that our theoretically grounded approach provides a general solution to the multiagent learning problem, which inherently comprises all key aspects of previous state of the art approaches on this topic. We test our method on a diverse suite of multiagent benchmarks and demonstrate a more efficient ability to adapt to new agents as they learn than baseline methods across the full spectrum of mixed incentive, competitive, and cooperative domains.
\end{abstract}

\input{introduction}
\input{problem_statement}
\input{approach}
\input{related_work}
\input{experiment}
\input{conclusion}

% Acknowledgements
\section*{Acknowledgements}
Research funded by IBM, Samsung (as part of the MIT-IBM Watson AI Lab initiative) and computational support through Amazon Web Services. Dong-Ki Kim was also supported by a Kwanjeong Educational Foundation Fellowship. The authors would like to thank anonymous reviewers for their helpful comments.

\bibliography{icml2021}
\bibliographystyle{icml2021}

\newpage
\appendix
\input{appendix}
\end{document}