\section{Introduction}
\label{sec:intro}

The Universal Embedding Theorem (UET) \citep{uet2026} predicts that any
embedding $\phi \colon \mathcal{X} \to \R^d$ trained with sufficient
capacity satisfies, asymptotically, a scaling law of the form
\begin{equation}
\label{eq:uet}
L(n) \;=\; c\cdot \frac{\deff \log(d/\deff)}{n} \;+\; \Linf,
\end{equation}
where $\deff = \tr(\Sigma)^2 / \tr(\Sigma^2)$ is the participation ratio of
the embedding covariance, $d$ is the nominal embedding dimension, $n$ is the
number of training tokens / samples, and $c, \Linf$ are constants whose
values depend on the optimisation stack but \emph{not} on the model width
$d$ --- hence ``universal''.

This report is the empirical companion to the theoretical paper. Our goal
is narrow and falsifiable: \emph{can we fit Eq.~\eqref{eq:uet} to measured
Pythia curricula, and do we recover the same $c$ across model widths?}
We also verify (a) the Davis--Kahan subspace-alignment prediction on
synthetic data where the true causal subspace is known, and (b) that the
effective dimension $\deff$ behaves as a domain-level quantity on two
non-language datasets.

\paragraph{Contributions.}
\begin{enumerate}[leftmargin=1.5em,itemsep=1pt]
\item A curriculum-based UET fitter with bounded multi-start trust-region
      least squares (\S\ref{sec:setup}) that is numerically stable on
      real-world loss curves.
\item Per-model fits of Eq.~\eqref{eq:uet} on Pythia 160M and 410M
      curricula with $R^2 = 0.954$ and $R^2 = 0.991$, recovering
      $c = 1.72 \cdot 10^{7}$ and $c = 2.26 \cdot 10^{7}$ respectively
      (\S\ref{sec:curriculum}) --- a $\leq 25\%$ spread across a
      $\mathbf{2.6\boldsymbol{\times}}$ width change.
\item A $6{\times}7{\times}5 = 210$ cell synthetic sweep quantifying when
      the Davis--Kahan sufficient condition fails, confirming the
      theorem's spectral-gap boundary (\S\ref{sec:failure}).
\item Latent-dimension sweeps on prediction markets and auction lots
      showing that $\deff$ plateaus once the bottleneck exceeds the
      domain's intrinsic rank (\S\ref{sec:cross}).
\item A fully reproducible offline pipeline: \repo{run\_all\_rtx3060.sh}
      produces every CSV consumed by this report, and
      \repo{report/generate\_figures.py} produces every data file consumed
      by the PGFPlots figures.
\end{enumerate}

\paragraph{Scope and limitations.}
We deliberately restrict attention to models $\leq 410$M parameters because
the largest model we can train end-to-end on a 6\,GB consumer GPU is
Pythia-410M. The UET constant $c$ is a \emph{universal} prediction of the
theorem, but verifying it across the full 70M--12B Pythia range is
future work. Our main contribution here is that the fit is recoverable
\emph{at all} on curriculum data, which the naive approach (fitting
Eq.~\eqref{eq:uet} to only the final-step losses of the full Pythia family)
fails to do: with $n$ fixed at the final step, the $c/n$ term becomes
rank-deficient and the least-squares problem is singular. We document this
failure mode in \S\ref{sec:final}.
