\section{Experimental Setup}
\label{sec:setup}

\subsection{Hardware and runtime}
All experiments run on a single Fedora 43 workstation:
Intel i9-11900H (16 threads), 64\,GB RAM, NVIDIA RTX~3060 Mobile (6\,GB
VRAM), CUDA 12.8, PyTorch 2.11. Total wall-clock for the full pipeline is
$\approx 75$ minutes. No cluster, no multi-GPU.

\subsection{Datasets}
\begin{itemize}[leftmargin=1.5em,itemsep=1pt]
\item \textbf{Pythia curricula.} We load the public EleutherAI/pythia-160m and
      pythia-410m checkpoints at 22 training steps each:
      \begin{center}\small
      $\{0, 1, 2, 4, 8, 16, 32, 64, 128, 256, 512,$\\
      $1000, 2000, 4000, 8000, 16000, 32000, 64000, 100000, 120000, 143000\}$.
      \end{center}
      At each checkpoint we measure validation
      loss on 140 Pile sequences and compute $\deff$ from the covariance of
      pooled hidden states.
\item \textbf{Pythia final checkpoints.} Four final-step checkpoints
      (70M, 160M, 410M, 1B) on the same evaluation set, used only to
      demonstrate the rank-deficient fit failure of \S\ref{sec:final}.
\item \textbf{Polymarket.} 40{,}470 prediction-market snapshots with 19
      structured features (price, volume, liquidity, category, ...).
\item \textbf{Auction catalogues.} 694{,}860 lots from Christie's and
      Sotheby's (\repo{silver\_extractions} tables) with 37 joined
      features.
\item \textbf{Synthetic Davis--Kahan sweep.} Grid over
      $d \in \{16,32,64,128,256,512\}$,
      $k \in \{2,4,8,16,32,64,128\}$,
      $\mathrm{gap} \in \{0.5,1,2,5,10\}$, with $10$ seeds per cell
      ($2{,}100$ rows total).
\end{itemize}

\subsection{Pythia tokens-per-step}
Pythia-deduped uses batch size $1024$ and sequence length $2048$, giving
$\text{tokens/step} = 1024 \cdot 2048 = 2{,}097{,}152$. We use this
conversion throughout:
\[
n(\text{step}) = \text{step} \cdot 2{,}097{,}152.
\]

\subsection{Effective dimension}
For a centred pooled embedding matrix $Z \in \R^{N\times d}$ with covariance
$\Sigma = Z^\top Z / N$, the effective dimension (participation ratio) is
\begin{equation}
\deff \;=\; \frac{\big(\tr \Sigma\big)^{2}}{\tr\big(\Sigma^{2}\big)}
         \;=\; \frac{\big(\sum_i \lambda_i\big)^2}{\sum_i \lambda_i^2},
\end{equation}
and the stable rank is $\|\Sigma\|_F^2 / \|\Sigma\|_2^2$. Both
quantities are dimensionless and invariant to overall scale.

\subsection{The UET fitter}
\label{sec:fitter}
We fit Eq.~\eqref{eq:uet} with a bounded multi-start trust-region
reflective least-squares solver (\texttt{scipy.optimize.least\_squares},
method \texttt{"trf"}). The two parameters are constrained to
$c \geq 0$, $0 \leq \Linf \leq 0.999 \min_j L_j$, which makes the problem
well-posed even at early training steps where loss is non-monotone.
For each of nine $\Linf$ initialisations in
$[0, 0.999\min_j L_j]$ we analytically solve the
$c$-initialisation by least-squares regression of $L - \Linf$ on the
feature $x = \deff\log(d/\deff)/n$, then run TRF to convergence and keep
the run with smallest RMSE. A synthetic recovery test
(\S\ref{sec:appendix-synth}) confirms the fitter recovers $c$ within
$10\%$ and $\Linf$ within $0.1$ of ground truth at realistic noise levels.

\subsection{Repository layout}
\begin{center}\small
\begin{tabular}{ll}
\toprule
Path & Purpose \\
\midrule
\repo{uet/scaling\_fit.py}     & UET fitter, token-conversion constants \\
\repo{uet/curriculum.py}       & Pythia checkpoint harvesting \\
\repo{uet/failure.py}          & Davis--Kahan synthetic sweep \\
\repo{uet/embedding\_train.py} & Bottleneck AE for domain data \\
\repo{scripts/run\_uet\_fit.py}& Fit driver, per-model + joint \\
\repo{run\_all\_rtx3060.sh}    & End-to-end pipeline (overnight) \\
\repo{report/generate\_figures.py} & CSV $\to$ PGFPlots \texttt{.dat} \\
\bottomrule
\end{tabular}
\end{center}
