\section{Discussion}
\label{sec:discussion}

\subsection{What we verified, what we did not}

\begin{center}\small
\begin{tabular}{p{5.5cm}p{6.0cm}l}
\toprule
Prediction & Evidence & Strength \\
\midrule
UET loss form fits a curriculum &
   Fig.~\ref{fig:uet-fit}, $R^2 \geq 0.95$ & strong \\
$c$ universal across model widths &
   Tab.~\ref{tab:uet-fit}, $\Delta c/c \leq 0.25$ over $2.5\times$ params
   & moderate \\
Spectral-gap clause is needed &
   Fig.~\ref{fig:failure-gap}, $4.3\times$ jump in $\sin\theta$ & strong \\
$k < d$ clause is needed &
   Fig.~\ref{fig:failure-d}, violations at small $d$ & strong \\
$\deff$ is domain intrinsic &
   Fig.~\ref{fig:cross-domain}, saturation at $\deff \approx 5$ & moderate \\
Discovery--Formalisation separation &
   Fig.~\ref{fig:curriculum-dynamics}, two regimes visible &
   qualitative \\
\bottomrule
\end{tabular}
\end{center}

\subsection{What breaks when you violate numerical assumptions}

The $R^2 \approx 0$ failure of \S\ref{sec:final} is the most important
diagnostic in this report. It is the standard trap of blindly feeding
loss numbers into a nonlinear fit: if the design matrix is
rank-deficient in \emph{any} coordinate, bounded optimisers will place
explanatory mass wherever the gradient is non-zero (here, entirely on
$\Linf$). The fix is not a better optimiser; it is a richer design
matrix, which curriculum sampling provides for free.

\subsection{Adam / AdamW learning-rate scaling}

Because every Pythia checkpoint was trained with the same optimiser
settings, we did not need to rescale $c$ for learning-rate changes.
If one were to combine checkpoints across optimiser regimes, the
Adam/AdamW square-root batch-size rule
\citep{loshchilov2019adamw}
\[
\eta_{\text{eff}} \;\propto\; \sqrt{\tfrac{B}{B_0}}
\]
becomes relevant: changing $B$ without rescaling $\eta$ shifts the
$c$ constant multiplicatively. A joint fit across different $B$ would
have to normalise by $\sqrt{B}$ first. We flag this as future work for
cross-family validations (e.g. Pythia + OLMo + Llama-Scaling-Suite).

\subsection{Compute honesty}

The entire empirical suite runs in $\approx 75$ minutes on a consumer
GPU and requires no cluster access. Extending to Pythia-1B curriculum
fitting (not just final-step evaluation) would require at most $6$
hours on an A100 and is the single highest-leverage follow-up. Nothing
in this report requires compute we do not have; nothing we claim
required more than what is available.

\subsection{What would falsify UET within our setup}

\begin{itemize}[leftmargin=1.5em,itemsep=2pt]
\item $R^2 < 0.8$ on either per-model curriculum fit.
\item $c$ values differing by more than a factor of $5$ between the
      two Pythia sizes.
\item $\sin\theta$ scaling unchanged by the spectral-gap clause in
      the synthetic sweep.
\item $\deff$ scaling linearly with the AE bottleneck (i.e.~no
      saturation).
\end{itemize}
None of these occurred.
